Donkeys to Monoids: A Computational Framework for Dynamic Semantics

### 2013-05-20 paper.tex
Addressed main points

\subsection{Donkey Monoids}

Imagine translating this sentence to DPL:
%
\begin{quotation}
A farmer owns a donkey, if he beats it
\end{quotation}
%
We would have to write something along the lines of:
%
\begin{equation}
\exists x \cdot farmer(x) \cdot \exists y \cdot donkey(y) \cdot beats(x,y) \rightarrow owns(x,y)
\end{equation}
%
The problem here is that reorderings like this work against our sought property of linearity: of being able to add new information as we go. We cannot compose `A farmer owns a donkey' and `if he beats it' in a meaningful way.

To rescue us from this problem Albert Visser suggests constructing a special e-monoid over the dynamic relational algebra. The idea is that we are going to simultaneously work on two `streams' of dpl, and switch easily between them with a special element $\Bowtie$.

Our sentence from above will then be expressible as
%
\begin{equation}
\Bowtie \cdot \exists x \cdot farmer(x) \cdot \exists y \cdot donkey(y) \cdot \Bowtie \cdot owns(x,y) \cdot \Bowtie \cdot beats(x,y)
\end{equation}
%
which can be written in `stream form' as
%
\begin{equation}
\begin{tabular}{|clll|}
    \hline
    $(-)$ & $\exists x \cdot farmer(x) \cdot \exists y \cdot donkey(y)$ & ~ & $beats(x,y)$ \\\hline
    $(+)$ & ~ & $owns(x,y)$ & ~ \\\hline
\end{tabular}
\end{equation}
%
The meaning of the above is $(-) \rightarrow (+)$

Another example from Visser is:
\begin{quotation}
A man keeps his word. He is honest
\end{quotation}
which roughly translates to:
\begin{equation}
\Bowtie\cdot\exists x\cdot man(x)\cdot\Bowtie\cdot keepword(x)\cdot honest(x)
\end{equation}
If you are interested in a lot more examples, see Visser's paper\cite{visser1999donkey}.

So how do we define this nifty $\Bowtie$? The trick is to move the entire dynamic relation algebra from before into an e-monoid that will help us keep track of some details.

If $\phi$ from before is the basic algebra, we specify the construction of our monoid $\Phi$ as follows.
%
\begin{flalign}
&\top := \langle \top, \top, + \rangle & \\
&\bot := \langle \top, \bot, + \rangle & \\
&\Bowtie := \langle \top, \top, - \rangle & \\
&\langle q_-, q_+, \alpha \rangle \cdot \langle r_-, r_+, \beta \rangle := \langle q_- \cdot r_{-\alpha}, q_+ \cdot r_{+\alpha}, \alpha\beta \rangle&
\end{flalign}
%
Notice how we keep the negative stream in the first position of the tuple, and the positive stream at the second position. The third position indicates to what stream we are currently writing. For example, if we are writing to the negative stream, and we apply an element that itself asks us to continue at its negative stream, we go back to the positive stream.

An even more interesting element than $\Bowtie$ is $\triangle$. $\triangle$ is a scope modifier, and it allows us to write:
%
\begin{quote}
A farmer owns a donkey. He beats it.
\end{quote}
%
(that is just some single farmer and donkey) as
%
\begin{equation}
\triangle \cdot \exists x \cdot farmer(x) \cdot \triangle \cdot owns(x,y) \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle \cdot beats(x,y)
\end{equation}
%
That is very close to the order of the natural language!

And combining the two monoids we can do our original sentence as
%
\begin{quote}
If a farmer owns a donkey, he beats it
\end{quote}
%
as
\begin{equation}
\Bowtie \cdot \triangle \cdot \exists x \cdot farmer(x) \cdot \triangle \cdot owns(x,y) \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle \cdot \Bowtie \cdot beats(x,y)
\end{equation}
%
In stream form, that is:
%
\begin{equation}
\begin{tabular}{|cllll|}
    \hline
    $(-,1)$ & $\exists x \cdot farmer(x)$ & ~ & ~ $ \exists y \cdot donkey(y)$ & ~ \\\hline
    $(-,0)$ & ~ & $owns(x,y)$ & ~ & ~ \\\hline
    $(+,1)$ & ~ & ~ & ~ & ~ \\\hline
    $(+,0)$ & ~ & ~ & ~ & $beats(x,y)$ \\\hline
\end{tabular}
\end{equation}
%
With the meaning being $(-,1) \cdot (-,0) \rightarrow (+,1) \cdot (+,0)$. Notice that the negative relations are again put on the left side of the implication, and the relations with larger scope value are put before those of lower value.

Notice $\Bowtie$ and $\triangle$ are commutative, since they work entirely on their own indicator. It's an interesting question to consider what sentences might require three or more levels of scoping. We could define $\triangle$ to go up in scope and $\bigtriangledown$ to go down. Instead of ${0,1}$ we would work on $\mathbb{Z}$.

For now we define the monoid as follows:
%
\begin{flalign}
&\top := \langle \top, \top, \top, \top, +, 0 \rangle & \\
&\bot := \langle \top, \top, \top, \bot, +, 0 \rangle & \\
&\Bowtie := \langle \top, \top, \top, \top -, 0 \rangle & \\
&\triangle := \langle \top, \top, \top, \top +, 1 \rangle & \\
&\langle q_{-,1}, q_{-,0}, q_{+,1}, q_{+,0}, \alpha, i \rangle \cdot \langle r_{-,1}, r_{-,0}, r_{+,1}, r_{+,0}, \beta, j \rangle := &\\
&\hspace{1cm}\langle q_{-,1} \cdot r_{-\alpha,1+i}, q_{-,0} \cdot r_{-\alpha,i}, q_{+,1} \cdot r_{\alpha,1+i}, q_{+,0} \cdot r_{\alpha,i}, \alpha\beta, i+j \rangle& \nonumber
\end{flalign}

In his paper Visser now goes on to define $\triangleleft$ and $\triangleright$ which work similarly to $\Bowtie$ except they work in different directions. 
%
\begin{equation}
\Bowtie \cdot \triangle \cdot \exists x \cdot farmer(x) \cdot \triangle \cdot owns(x,y) \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle \cdot \triangleleft \cdot beats(x,y)
\end{equation}
%
This is quite interesting for certain sentences containing `surprises', like
%
\begin{quotation}
He sees... no donkey
\end{quotation}
%
In this case we would compositionally assume to start out with
$sees(x,y)$ and finish with $\cdot \exists y \cdot donkey(y)\triangle$ hence we want to write something like:
%
\begin{equation}
sees(x,y) \cdot \triangleleft \cdot \bot \cdot \triangleright \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle
\end{equation}
%
Notice that that means the word `no'/`not' must be $\triangleleft\cdot\bot\cdot\triangleright$. Unfortunately these retrospective triangles tend to mess sentence enough up, that they require lots of `stream unification brackets', and don't generally give very nice formulas. In the semantic parsing in the next section, they wont be used at all.

@@

Makoto Kanazawa has written an entire paper on issues with weak and strong readings of donkey sentences\cite{kanazawa1994weak}. For the donkey sentence, we may have at least the following readings of the second `a':
\begin{enumerate}
\item Weak reading: A farmer who owns a donkey beats a donkey he owns. 
\item Strong reading: A farmer who owns a donkey beats every donkey he owns 
\item Unique reading: A farmer who owns a donkey beats the donkey he owns. 
\end{enumerate}

@@

In conclusion, Montague like grammars for dependency graphs could be very interesting to make. It would however require some well defined rules for what we can expect to find under each type of arrow. This was a blind road.

@@

This means that nearly all of the rules I shall present are of my own creation. Because these rules are tested in practice on a large set of English sentences, they will hopefully avoid some simple mistakes and oversights that are easy to make in purely theoretical settings. It seams like a good approach to start from a small working fragment, and slowly add more phrase rules and sentences. The opposite of this approach is what is often observed when logicians and philosophers go astray in very abstract corners of English, without having covered the basic ground.

>>>>

\subsection{Donkey Monoids}
With the introduction of DPL we came a long way in terms of a compositional logic. However looking once more at the donkey sentence variation:
%
\begin{quotation}
A farmer owns a donkey, if he beats it.
\end{quotation}
%
And its DPL formula:
%
\begin{equation}
\exists x \cdot farmer(x) \cdot \exists y \cdot donkey(y) \cdot beats(x,y) \rightarrow owns(x,y)\label{eq:dpl_beats_if_owns}
\end{equation}
%
We see that we still haven't reached the compositional Nirvana. Equation \eqref{eq:dpl_beats_if_owns} is clearly not based on the composition of `A farmer owns a donkey' and `if he beats it' in a direct way.

\subsubsection{Polarity Switcher}
To help our course, Albert Visser suggests the construction of a special monoid over DPL. The idea is that we are going to simultaneously work on two `streams' of DPL, and switch easily between them with a special constant $\Bowtie$.

The inverse donkey sentence can now be expressed as:
%
\begin{equation}
\Bowtie \cdot \exists x \cdot farmer(x) \cdot \exists y \cdot donkey(y) \cdot \Bowtie \cdot owns(x,y) \cdot \Bowtie \cdot beats(x,y)
\end{equation}
%
which `internally' is represented by the following streams of information:
%
\begin{equation}
\begin{tabular}{|clll|}
    \hline
    $(-)$ & $\exists x \cdot farmer(x) \cdot \exists y \cdot donkey(y)$ & ~ & $beats(x,y)$ \\\hline
    $(+)$ & ~ & $owns(x,y)$ & ~ \\\hline
\end{tabular}
\end{equation}
%
Visser interprets the above as the DPL formula $(-) \rightarrow (+)$.

Another example from Visser is: `A man keeps his word. He is honest' which roughly translates into: $\Bowtie\cdot\exists x\cdot man(x)\cdot\Bowtie\cdot keepword(x)\cdot honest(x)$ for the general reading. Visser has a whole lot more examples in his paper\cite{visser1999donkey}.

So how do we define this nifty $\Bowtie$? The trick is to move the entire dynamic predicate logic from before into an monoid that will help us keep track of some extra details.

A monoid is an simple algebraic structure with a one associative binary operation and an identity element. Our monoid $\langle\mathcal{D},\cdot\rangle$ is going to have the domain $\mathcal{D}=DPL\times DPL\times\{+,-\}$, which is populated by:
\begin{equation}
\langle\top,\phi,+\rangle \in \mathcal{D} \quad \text{for all}\enspace \phi\in\text{DPL}
\end{equation}
With no chance of confusion, we will simply denote these elements by their DPL representation. For example $\mathcal{D}$ contains the constants $\top$ and $\bot$ and the $\Bowtie$ constant here defined:
\begin{align}
\top &:= \langle \top, \top, + \rangle \\
\bot &:= \langle \top, \bot, + \rangle \\
\Bowtie &:= \langle \top, \top, - \rangle
\end{align}
The associative operation $\cdot$ is as below. To make the definition slicker, we give the set $\{+,-\}$ a multiplication operation $++\mapsto+$, $+-\mapsto-$, $-+\mapsto-$, $--\mapsto+$.
\begin{align}
&\langle q_-, q_+, \alpha \rangle \cdot \langle r_-, r_+, \beta \rangle := \langle q_- \cdot r_{-\alpha},\enspace q_+ \cdot r_{+\alpha},\enspace  \alpha\beta \rangle
\end{align}
%
Notice how we keep the negative stream in the first position of the tuple, and the positive stream at the second position. The third position indicates to what stream we are currently writing. For example, if we are writing to the negative stream, and we apply an element that itself asks us to continue at its negative stream, we go back to the positive stream.

\subsubsection{Scope Modifier}
We won't spend much time with the above monoid though, since we are immediately going to introduce a generalization! We are going to introduce the scope modifier $\triangle$, which allows us to write:
%
\begin{quote}
A farmer owns a donkey. He beats it.
\end{quote}
%
(that is just some single farmer and donkey) as
%
\begin{equation}
\exists x \cdot farmer(x) \cdot owns(x,y) \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle \cdot beats(x,y)
\end{equation}
%
That is very close to the order of the natural language!

And combining the scope modifier with the polarity switched we can finally, as promised, write our original sentence `If a farmer owns a donkey, he beats it' as:
\begin{equation}
\Bowtie \cdot \exists x \cdot farmer(x) \cdot owns(x,y) \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle \cdot \Bowtie \cdot beats(x,y)
\end{equation}
%
In stream form, that is:
%
\begin{equation}
\begin{tabular}{|cllll|}
    \hline
    $(-,1)$ & ~ & ~ & ~ $ \exists y \cdot donkey(y)$ & ~ \\\hline
    $(-,0)$ & $\exists x \cdot farmer(x)$ & $owns(x,y)$ & ~ & ~ \\\hline
    $(+,1)$ & ~ & ~ & ~ & ~ \\\hline
    $(+,0)$ & ~ & ~ & ~ & $beats(x,y)$ \\\hline
\end{tabular}
\end{equation}
%
With the DPL interpretation being $(-,1) \cdot (-,0) \rightarrow (+,1) \cdot (+,0)$. Notice that the negative relations are again put on the left side of the implication, and the relations with larger scope value are put before those of lower value. We see $\Bowtie$ and $\triangle$ are commutative, since they work entirely on their own indicator.

As expected from the above diagram, we are going to define our new monoid $\langle\mathcal{D},\cdot\rangle$ as having the domain\footnote{We have chosen to model our scopes in $\mathbb{Z}_2$. It's an interesting question to consider what sentences might require three or more levels of scoping. If we chose to work on the entire $\mathbb{Z}$, we could define $\triangle$ to go up in scope and $\triangledown$ to go down.} $DPL^4\times\{+,-\}\times\mathbb{Z}_2$ with the elements:
\begin{equation}
\langle \top, \top, \top, \phi, +, 0 \rangle \in \mathcal{D} \quad \text{for all}\enspace \phi\in\text{DPL}
\end{equation}
In addition to $\Bowtie$ and $\triangle$ defined as:
\begin{align}
\top &:= \langle \top, \top, \top, \top, +, 0 \rangle \\
\bot &:= \langle \top, \top, \top, \bot, +, 0 \rangle \\
\Bowtie &:= \langle \top, \top, \top, \top -, 0 \rangle \\
\triangle &:= \langle \top, \top, \top, \top +, 1 \rangle
\end{align}
For $\cdot$ the definition works just like in the previous case, though slightly more complicated:
\begin{align}
&\langle q_{-,1}, q_{-,0}, q_{+,1}, q_{+,0}, \alpha, i \rangle \cdot \langle r_{-,1}, r_{-,0}, r_{+,1}, r_{+,0}, \beta, j \rangle := &\\
&\hspace{1cm}\langle q_{-,1} \cdot r_{-\alpha,1+i},\enspace q_{-,0} \cdot r_{-\alpha,i},\enspace q_{+,1} \cdot r_{\alpha,1+i},\enspace q_{+,0} \cdot r_{\alpha,i},\enspace \alpha\beta,\enspace i+j \rangle& \nonumber
\end{align}

In his paper Visser now goes on to define $\triangleleft$ and $\triangleright$ which work similarly to $\Bowtie$ except they work `backwards and forward in time'. This allows us to model `if he beats it' entirely on its own as $\triangleleft\cdot beats(x,y)$ in
\begin{equation}
  \exists x \cdot farmer(x) \cdot owns(x,y) \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle \cdot \triangleleft \cdot beats(x,y)
\end{equation}
%
If we immagined a system that tried to build meaning from its input as soon as it came in, $\triangleleft$ would allow us to account for `surprises' like in the sentence:
\begin{quotation}
He sees... no donkey
\end{quotation}
%
In this case we would compositionally like to start out with $sees(x,y)$ expecting a $\triangle\cdot\exists y\cdot\triangle$ somehow, but when we instead see the negation, we end up with the formula:
%
\begin{equation}
sees(x,y) \cdot \triangleleft \cdot \bot \cdot \triangleright \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle
\end{equation}
%
Visser suggests $\triangleleft\cdot\bot\cdot\triangleright$ often is a good candidate for words like `no' and `nobody'. Unfortunately these retrospective triangles tend to mess sentences enough up, that I had to include brackets everywhere to keep things under control. More brackets are seldom the way to beautiful formulas, so in the automatic framework in the next section, we will settle with the $\Bowtie$ and $\triangle$ monoid.

@@

Makoto Kanazawa has written an entire paper on issues with weak and strong readings of donkey sentences\cite{kanazawa1994weak}. For our main donkey sentence, we may have at least the following readings of the second `a':
\begin{enumerate}
\item Weak reading: A farmer who owns a donkey beats a donkey he owns. 
\item Strong reading: A farmer who owns a donkey beats every donkey he owns 
\item Unique reading: A farmer who owns a donkey beats the donkey he owns. 
\end{enumerate}

@@

In conclusion, Montague like grammars for dependency graphs could be very interesting to make. It would however require some well defined rules for what we can expect to find under each type of arrow. This was a blind alley.

@@

This means that nearly all of the rules I shall present are of my own creation. Because these rules are tested in practice on a large set of English sentences, they will hopefully avoid some simple mistakes and oversights that are easy to make in purely theoretical settings. It seems like a good approach to start from a small working fragment, and slowly add more phrase rules and sentences. The opposite of this approach is what is often observed when logicians and philosophers go astray in very abstract corners of English, without having covered the basic ground.
